Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-18 paper.tex
...

>>>>

\subsection{Dispatching Hierarchies \\ Case Study: Bomber `draw` and `update` callbacks}

Entities typically form a dispatching hierarchy in which a container that
receives a stimulus automatically forwards it to its managed children.

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
class Bomber : public PinguAction {
    <...>
    Sprite sprite;
}

Bomber::Bomber (<...>) : <...> {
    sprite.load(<...>);
    <...>
}

void Bomber::update () {
    sprite.update ();
}

void Bomber::draw (SceneContext& gc) {
    <...>
    gc.color().draw(sprite, <...>);
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName do
    <...>
    var&? Sprite sprite = spawn Sprite(<...>);
    <...>
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{hierarchy}
\caption{Dispatching chain for \code{update}.
\label{fig.hier}
}
\end{figure}

In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
animation frames (Figure~\ref{fig.TODO}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus from the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen.
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj}.
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus.
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection.
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions.
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

In \CEU, the \code{Bomber} action spawns a \code{Sprite} animation instance on
its body.
%
The \code{Sprite} instance (ln. 3) can react directly to external \code{dt} and
\code{redraw} events (which are analogous to \code{update} and \code{redraw}
callbacks, respectively), bypassing the program hierarchy entirely.
While and \emph{only while} the bomber abstraction is alive, the sprite
animation is also alive.
The radical decoupling between the program hierarchy and reactions to events
eliminates dispatching chains entirely.

\subsection{Lifespan Hierarchies \\ Case Study: The Pingus Container}

Entities typically form a lifespan hierarchy in which a terminating
container entity automatically destroys its managed children.

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Pingu* PinguHolder::create_pingu (<...>) {
    <...>
    Pingu* pingu = new Pingu (<...>);
    pingus.push_back(pingu);
    <...>
}

void PinguHolder::update() {
    <...>
    while(pingu != pingus.end()) {
        (*pingu)->update();
        if ((*pingu)->get_status() == Pingu::PS_DEAD) {
            pingu = pingus.erase(pingu);
        }
        <...>
        ++pingu;
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.40\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Game (void) do
    <...>
    pool[] Pingu pingus;
    code/await Pingu_Spawn (<...>) do
        <...>
        spawn Pingu(<...>) in outer.pingus;
    end
    <...>   // code invoking Pingu_Spawn
end

TODO

code/await Pingu (<...>) do
    <...>
    loop do
        await outer.game.dt;
        if call Pingu_Rel_Getpixel(0,-1) == {Groundtype::GP_OUTOFSCREEN} then
            <...>
            escape {PS_DEAD};
        end
    end
end
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

In C++, the class \code{PinguHolder} is a container that holds all pingus
alive.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection with updates on every frame (ln. 11).
Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
go unnoticed but the program would keep consuming memory and CPU time.
This problem is known as the \emph{lapsed listener}~\cite{TODO} and also occurs
in languages with garbage collection:
A container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{pool}
\caption{Lifespan of dynamic instances.
\label{fig.pool}
}
\end{figure}

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen spawns a new \code{Pingu} on every invocation of
\code{Pingu\_Spawn}.
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its *natural termination*, the instance is automatically removed
from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. TODO) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

\subsection{Signaling Mechanisms\\ Case Study: Global Keys and the Options Menu}

Entities often need to communicate explicitly through signaling mechanisms,
especially if there is no hierarchy relationship between them.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{The \emph{Mouse Grab} configuration option.
\label{fig.options}
}
\end{figure}

The *Mouse Grab* option restricts the mouse movement to the game window
boundaries (Figure~\ref{fig.options}).
The option can be set anywhere in the game by pressing *Ctrl-G*.
In addition, the *Options* menu has a check box to toggle the *Mouse Grab*
option with mouse clicks while still responding to *Ctrl-G* presses.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{options-anim-opt}
\caption{Mutual dependency between signals.
\label{fig.events}
}
\end{figure}

The implementations in C++ and \CEU use a signalling mechanism to connect the
key presses, the check box, and a configuration manager that applies the
appropriate side effects in the game (i.e., restrict the mouse movement).
Figure~\ref{fig.events} illustrates how the mutual notifications create a 
dependency cycle between the configuration manager and the check box.

\begin{figure*}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
void GlobalEvent::on_button_press (<...>) {
    <...>
    switch (event.keysym.sym) {
        case SDLK_g:
            if (keystate[SDLK_LCTRL] || keystate[SDLK_RCTRL]) {
                config_manager.set_mouse_grab(          @ctrl_g_cpp
                    !config_manager.get_mouse_grab());
            }
            break;
        <...>
    }
}

///

boost::signals2::signal<void(bool)> on_mouse_grab_change;   // definition in `config_manager.h` @signal_def

void ConfigManager::set_mouse_grab (bool v) {   @set_mouse_grab
    <...>
    if (v != get_mouse_grab()) {                @if_1
        <...>   // the actual "grab" effect
        on_mouse_grab_change(v);                @signal
    }                                           @if_2
}

///

boost::signals2::signal<void (bool)> on_change;   // definition in `check_box.hpp`

void CheckBox::set_state (bool is_on, bool send_signal) {   @last_argument
    <...>   // switches the check box state
    if (send_signal) {      @if_cb_1
        on_change(is_on);
    }                       @if_cb_2
}

///

typedef std::vector&lt;boost::signals2::connection&gt; Connections;   // definition in `option_menu.hpp`
Connections connections;                                        // definition in `option_menu.hpp`

OptionMenu::OptionMenu() :
    connections(),
    mousegrab_box(),
    <...>
{
    mousegrab_box = new CheckBox(<...>);
    connections.push_back(                              @bind_11
        config_manager.on_mouse_grab_change.connect(
            std::bind(&CheckBox::set_state, mousegrab_box, <...>, false); @bind_false
        )
    );                                                  @bind_12
    connections.push_back(                              @bind_21
        mousegrab_box->on_change.connect(
            std::bind(&ConfigManager::set_mouse_grab, &config_manager, <...>);
        )
    );                                                  @bind_22
    <...>

}

OptionMenu::~OptionMenu() {     @destr_1
    for (Connections::iterator i=connections.begin(); i!=connections.end(); ++i) {
        (*i).disconnect();
    }
}                               @destr_2

\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.40\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
spawn do
    var _SDL_KeyboardEvent&& e;
    every e in SDL_KEYDOWN do
        var _u8&& keystate = _SDL_GetKeyState(null);
        <...>
        if e:keysym.sym == {SDLK_g} then
            if ((keystate[{SDLK_LCTRL}] as bool) or (keystate[{SDLK_RCTRL}] as bool)) then
                emit config_manager.go_mouse_grab(    @ctrl_g_ceu
                        not ({config_manager.get_mouse_grab()} as bool));
            end
        end
        <...>
    end
end

///

data ConfigManager with
    event bool go_mouse_grab;
end
var ConfigManager config_manager = val ConfigManager(_);

spawn do
    var bool v;
    every v in config_manager.go_mouse_grab do
        <...>   // the actual "grab" effect
    end
end

///

data ICheckBox with
    var   bool is_on;
    event bool go_click;
end

code/await CheckBox (<...>) -> (var ICheckBox checkbox) -> FOREVER do
    checkbox = val ICheckBox(<...>);
    <...>
    par do
        every c.component.on_click do                   @every_1
            emit checkbox.go_click(not checkbox.is_on); @dir_class_app
        end                                             @every_2
    with
        loop do                                         @loop_1
            <...>   // switches the check box state
            checkbox.is_on = await checkbox.go_click;   @dir_app_class
        end                                             @loop_2
    end
end

///

code/await OptionMenu <...> do
    <...>
    var& CheckBox b2 = <...>;
    spawn do
        par do
            var bool v;
            every v in outer.config_manager.go_mouse_grab do    @loop_11
                emit b2.checkbox.go_click(v);
            end                                                 @loop_12
        with
            var bool v;
            every v in b2.checkbox.go_click do                  @loop_21
                emit outer.config_manager.go_mouse_grab(v);
            end                                                 @loop_22
        end
    end
    <...>
end

\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

In C++, the class `GlobalEvent` detects *Ctrl-G*
presses and invokes the callback `config\_manager.set\_mouse\_grab`:
%
The class `ConfigManager` uses a `boost::signal`
[[![X]][boost\_signal]] to notify the application when the new configuration is
applied:
%
The `if` enclosing the signal emission @NN(if\_1,-,if\_2) breaks the dependency 
cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
%
The class `CheckBox` also uses a `boost::signal` to 
notify the application on changes:
%
Again, the `if` enclosing the signal emission @NN(if\_cb\_1,-,if\_cb\_2) breaks the 
dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
%
The class `OptionMenu` creates the dependency loop by
connecting the two signals:
%
The constructor binds
the signal `config\_manager.on\_mouse\_grab\_change` to the callback method
           `mousegrab\_box->set\_state`
           @NN(bind\_11,-,bind\_12),
and also
the signal `mousegrab\_box->on\_change` to the callback method
           `config\_manager.set\_mouse\_grab`
           @NN(bind\_21,-,bind\_22).
This way, every time the `ConfigManager` signals `on\_mouse\_grab\_change`
(`ConfigManager`, ln. @N(signal) [up](\#cpp\_config-manager)), `set\_state` is
implicitly called.
The same happens between the signal `on\_change` in the `CheckBox` and the 
method `set\_mouse\_grab` in the `ConfigManager`
(`ConfigManager`, ln. @N(set\_mouse\_grab) [up](\#cpp\_config-manager)).
%
Note that the signal binding to call `CheckBox::set\_state` @NN(bind\_false) 
receives a fixed value `false` as the last argument to prevent infinite
execution (`CheckBox`, ln. @N(last\_argument) [up](\#cpp\_check-box)).
%
The destructor @NN(destr\_1,-,destr\_2) breaks the connections explicitly when
the *Option* screen terminates.

In \CEU, a *Ctrl-G* key press broadcasts the internal event
`config\_manager.go\_mouse\_grab` to the application [[![X]][ceu\_global\_event]]:
%
The configuration manager [[![X]][ceu\_config\_manager]] just needs to react to
`go\_mouse\_grab` continuously to perform the *grab* effect:
%
The `CheckBox` [[![X]][ceu\_check\_box]] exposes the event `go\_click` for 
notifications in both directions, i.e., from the abstraction to the application
and *vice versa*:
%
The abstraction reacts to external clicks continuously @NN(every\_1,-,every\_2)
to broadcast the event `go\_click` @NN(dir\_class\_app).
It also reacts continuously to `go\_click` in another line of execution
@NN(loop\_1,-,loop\_2), which awakes from notifications from the first line of
execution or from the application.
%
The `OptionMenu` [[![X]][ceu\_option\_menu]] connects the two events as 
follows:
%
The two loops in parallel handle the connections in opposite directions:
from the configuration manager to the check box @NN(loop\_11,-,loop\_12);
and
from the check box to the configuration manager @NN(loop\_21,-,loop\_22).
%
When the *Option* screen terminates, the connections break automatically since
the body is automatically aborted.
%
Note that the implementation in \CEU does not check event emits to break the
dependency cycle and prevent infinite execution.
Due to the [stack-based execution for internal events][ceu\_stack] in \CEU,
programs with mutually-dependent events do not create infinite execution loops.

\section{Related Work}

\section{Conclusion}

We promote the \emph{structured synchronous reactive} programming model of
\CEU for the development of games.
We present in-depth use cases categorized in 5 control-flow patterns applied to
\emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to other
games.

We show how the standard way to program games with objects and callbacks in C++
hinders structured programming techniques, such as support for sequential
execution, long-lasting loops, and persisting local variables.
In this sense, callbacks actually disrupt structured programming, becoming
["our generation’s goto"][goto] according to Miguel de Icaza.

Overall, we believe that most difficulties in implementing control behavior in 
game logic is not inherent to this domain, but a result of accidental
complexity due to the lack of structured abstractions and an appropriate
concurrency model to handle event-based applications.

[goto]: http://tirania.org/blog/archive/2013/Aug-15.html

\section{Acknowledgments}

[Leonardo Kaplan](https://github.com/leokaplan/) and
[Alexander Tkachov](https://github.com/Tkachov/)
for early explorations and prototypes.
